---
title: "Student Dashboard"
subtitle: "Integrated Python + R Analytics (Supabase)"
author: "Kyle Miranda"
date: last-modified
format: 
  html:
    embed-resources: true
    toc: true
    toc-location: left
    theme: cosmo
    code-fold: true
    code-summary: "View Code"
execute:
  echo: false
  warning: false
  message: false
---

> **Portfolio Note:** This report demonstrates a polyglot data workflow. I utilized **Python** for its robust API ecosystem to interface with Supabase and perform complex data merging. I then leveraged **R** and the `reticulate` library to generate publication-quality tables and statistical visualizations, showcasing the ability to choose the best tool for each stage of the data lifecycle.

```{r setup, include=FALSE}
#| label: load-bridge
# This R chunk loads the bridge between Python and R
library(reticulate)
library(tidyverse)
library(gt)
library(patchwork)
```

```{python, include=FALSE, echo=FALSE}
# ─── IMPORT REQUIRED PACKAGES ──────────────────────────────
#| label: python-extraction
# ─── DATA ENGINEERING (PYTHON) ──────────────────────────────
import pandas as pd
import numpy as np

# Mocking the import for demonstration - keep your 'from main import...' line
from main import students_df, attendance_df, assessments_df, schools_df

# Merge logic (Efficient in Python/Pandas)
assessment_merged = assessments_df.merge(students_df, on='student_id', how='left')
assessment_merged = assessment_merged.merge(schools_df, on='school_id', how='left')

assessment_merged['EL_flag'] = assessment_merged['ell_status'].str.lower().map({'yes': 'EL', 'no': 'Non-EL'})
assessment_merged['SPED_flag'] = assessment_merged['special_ed_flag'].str.lower().map({'yes': 'SPED', 'no': 'Non-SPED'})

print("✅ Data Loaded and Merged")
```

-   [Overview](#overview)
-   [Student Distribution](#student-distribution)
-   [Attendance Analysis](#attendance-analysis)
-   [Assessment Analysis](#assessment-analysis)
-   [School Comparison](#school-comparison)
-   [Interactive Tools](#interactive-tools)
-   [Correlation Analysis](#correlation-analysis)

## Overview {#overview}

```{r}
#| label: overview-table
library(reticulate)
library(tidyverse)
library(gt)

students <- py$students_df
assessments <- py$assessments_df

total_students <- nrow(students)
avg_score <- mean(assessments$score)
ell_pct <- mean(students$ell_status == "Yes") * 100
sped_pct <- mean(students$special_ed_flag == "Yes") * 100

tibble(
  Metric = c("Total Students","Average Score","% EL","% SPED"),
  Value = c(total_students, round(avg_score,1), paste0(round(ell_pct,1),"%"), paste0(round(sped_pct,1),"%"))
) %>%
  gt() %>%
  tab_header(title = "Key Metrics") %>%
  cols_align("center") %>%
  tab_style(
    style = cell_text(weight = "bold", size = px(18)),
    locations = cells_body(columns = Value)
  )
```

## Student Distribution {#student-distribution}
```{r}
#| label: Student distribution by grade
# Student count by grade and gender
py$students_df %>%
  count(grade_level) %>%
  ggplot(aes(x=factor(grade_level), y=n)) +
  geom_col(fill="#2C7FB8") +
  labs(title="Student Distribution by Grade", x="Grade", y="Count") +
  theme_minimal()

```

## Attendance Analysis {#attendance-analysis}
```{r}
#| label: attendance-viz
#| echo: false
# Process the python attendance dataframe in R
py$attendance_df %>%
  left_join(py$students_df, by="student_id") %>%
  count(grade_level, status) %>%
  ggplot(aes(x=factor(grade_level), y=n, fill=status)) +
  geom_col(position="dodge") +
  scale_fill_manual(values=c("Absent"="#d73027","Tardy"="#fc8d59","Present"="#1a9850")) +
  labs(title="Attendance by Grade", x="Grade", y="Days") +
  theme_minimal()

```


## Assessment Analysis {#assessment-analysis}
### EL vs Non-EL
```{r}
#| label: assessment-boxplots
ggplot(py$assessment_merged, aes(x=factor(grade_level), y=score, fill=EL_flag)) +
  geom_boxplot() +
  facet_wrap(~subject) +
  scale_fill_manual(values=c("EL"="#d73027","Non-EL"="#1a9850")) +
  labs(title="Score Distribution (EL vs Non-EL)", x="Grade", y="Score") +
  theme_minimal()
```

### SPED vs Non-SPED
```{r}
  ggplot(py$assessment_merged, aes(x=factor(grade_level), y=score, fill=SPED_flag)) +
  geom_boxplot() +
  facet_wrap(~subject) +
  scale_fill_manual(values=c("SPED"="#756bb1","Non-SPED"="#31a354")) +
  labs(title="Score Distribution (SPED vs Non-SPED)", x="Grade", y="Score") +
  theme_minimal()
```

## School Comparison {#school-comparison}
```{python}
#| label: Average assessment scores by school
# Merge assessments with school info
bins = [0,50,60,70,80,90,100]
labels = ['0-50','51-60','61-70','71-80','81-90','91-100']

assessment_merged['score_bin'] = pd.cut(assessment_merged['score'], bins=bins, labels=labels)

school_bins = assessment_merged.groupby(['school_name','score_bin']).size().reset_index(name='count')
```

```{r}
ggplot(py$school_bins, aes(x=school_name, y=count, fill=score_bin)) +
  geom_col() +
  labs(title="Score Distribution by School", x="School", y="Students") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Correlation Analysis {#correlation-analysis}
```{python}
#| label: python-correlation
#| echo: true
# Demonstrating Python's strength in matrix manipulation
# Encode categorical → numeric
corr_df = assessment_merged.copy()

corr_df['EL_num'] = corr_df['EL_flag'].map({'EL':1,'Non-EL':0})
corr_df['SPED_num'] = corr_df['SPED_flag'].map({'SPED':1,'Non-SPED':0})

pivot_scores = corr_df.pivot_table(index='student_id', columns='subject', values='score')

final_corr = pivot_scores.copy()
final_corr['EL'] = corr_df.groupby('student_id')['EL_num'].first()
final_corr['SPED'] = corr_df.groupby('student_id')['SPED_num'].first()
final_corr['Grade'] = corr_df.groupby('student_id')['grade_level'].first()

corr_matrix = final_corr.corr()
```


```{r}
#| label: r-heatmap
# Visualizing the Python matrix using R's aesthetaic control
library(corrplot)

# Convert the Python object to an R Matrix explicitly
m <- as.matrix(py$corr_matrix)

corrplot(m,
         method = "color",
         addCoef.col = "black",
         tl.col = "black",
         number.cex = 0.7,
         # Add a little top margin so the title doesn't get cut off
         mar = c(0, 0, 2, 0), 
         title = "Correlation Matrix: Scores, EL, SPED, Grade")

```